---
title: "DeepSig: Machine Learning-based Mutational Signature Inference for WES and MSK-IMPACT  data"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{DeepSig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# De novo inference
```{r load, eval=TRUE}
library(DeepSig)
```

We illustrate de novo inferences using TCGA-BRCA catalog data:
```{r tcga-brca, eval=TRUE}
brca <- read.table(system.file('extdata', 'tcga_brca_mutect_catalog.txt',
                               package = 'DeepSig'))
tcga <- DeepSig(data = brca)
tcga
```

## Number of signatures
`DeepSig` provides a way to determine the most likely number $K$ of signatures present 
using a Bayesian marginal likelihood method. To do this, scan a range of $K$ values via
```{r bnmf0, echo = TRUE, eval = FALSE}
set.seed(152)
tscan <- extractSig(tcga, method = 'bnmf', Kmin = 2, Kmax = 10, nrun = 10, progress.bar = TRUE)
```

```{r bnmf, echo = FALSE, eval = TRUE}
set.seed(152)
tscan <- extractSig(tcga, method = 'bnmf', Kmin = 2, Kmax = 10, nrun = 10, progress.bar = FALSE)
```

```{r bnmf1, fig.width=5, fig.height=4, fig.fullwidth = FALSE}
me <- misc(tscan)$measure
me
plot(me, type = 'b', las = 1, bty = 'n')
```

Note that individual runs consist of $K$ values scanned from `Kmin` (default 2) to `Kmax`. 
Often, increasing the size of signature set by one will result in redundant signatures beyond a certain $K$ value. If this occurrs, the scan will stop and print this upper bound `Kmax`. 
In the example above, this upper bound is often smaller than the maximum 10.

## Decomposition under optimal K
The LML (log marginal likelihood) plot above suggests a minimal $K = 5$.
It is then recommended to re-run the de novo inference using `method = "nmf"`
under `K = 5`:

```{r bnmf2, echo = TRUE, eval = FALSE}
tcga <- extractSig(tcga, method = 'nmf', K = 5, nrun = 100, progress.bar = TRUE)
sig <- signat(tcga)
head(sig)
```
```{r bnmf2_2, echo = FALSE, eval = TRUE}
tcga <- extractSig(tcga, method = 'nmf', K = 5, nrun = 100, progress.bar = FALSE)
sig <- signat(tcga)
head(sig)
```

## Signature annotation
The main object of interest within `tcga` is the de novo signature set that
can be extracted as `signat(tcga)`. We compute a cosine similarity matrix
between de novo and reference signatures (COSMIC v3)
```{r bnmf3, eval = TRUE}
cs <- denovoCosim(sig = sig, ref.sig = 'cosmic')
cs[,1:5]
```

There is a customized heatmap for displaying `cs`:
```{r bnmf}, fig.width=10, fig.height=10, fig.fullwidth = TRUE}
csHeatmap(cs)
```

In the above, we computed the cosine similarity matrix of the 5 de novo signatures with the COSMIC v3 set.
The heatmap indicates that `S1-S5` correspond to SBS3 (HRD), SBS10 (POLE), SBS6 (MMR), SBS1(aging), and
SBS2.13 (APOBEC), respectively. To faciliate further analyses, it is
convenient to re-label de novo signatures based on these annotations:
```{r lsap, eval = TRUE}
annotation <- c('SBS_1'='S4','SBS_2.13'='S5','SBS_3'='S1','SBS_6'='S3','SBS_10'='S2')
tcga2 <- reorderSig(tcga, annotation = annotation)
head(signat(tcga2))
```
The function `reorderSig` reorders the signatures of the object in the sequence of
`annotation` parameter and renames them according to `names(annotation)`.

We now examine the actual shape simlarity of de novo and reference signatures:
```{r sigplot, eval = TRUE, fig.width=10, fig.height=10, fig.fullwidth = TRUE}
denovoProfiles(tcga2, ref.sig = 'cosmic')
```
The left-most column shows de novo signatures and the second and third columns 
show the reference signatures with the top and second-highest cosine similarity
overlaps (in parentheses) with de novo signatures. Note that `SBS_2` is in 
fact a composite of `SBS2` and `SBS13`, which is a very common occurence; 
the two APOBEC signatures rarely are seen separately in non-WGS samples. The 
same is true for `SBS_10`, which is a composite of `SBS10a` and `SBS10b`.